Deadlock Freedom on Self-Assembled DNA Nano Networks

### 2013-11-04 trunk/PAPER/disr.tex
Synced to after-conferences updates

\subsection{Basic Idea}
The main concept behind of the original SR approach is the
\emph{Segment}: a Segment is basically a path of consecutive node
and links. All segments start with a link attached to a node which
belongs to different segment, ending with a link also attached to a
node belonging to a different segment. A exception is the first
segment estabilished in the network, called \emph{starting segment} which is  a loop
beginning/ending on a particular node defined as bootstrap node.

The idea is to partition the network in a set of disjoint segments, and then
placing a turn restriction within each segment. It has been proved~\cite{mejia_ipdps06}
that such a set of turn prohibitions guarantees dealock freedom and
while preserving connectivity of the network.

@@

Roughly speaking, the execution of can be described as the execution
of the following phases:
\begin{itemize}
\item Injection of the DiSR process from upper layer to set a bootstrap
node
\item Bootstrap node broadcasting to create the first segment of the subnet
\item Parallel requests starting from assigned node to discover other segments
%\item Turn proibition in each confirmed segment
%\item Draining of remaining packet due timeouts
\end{itemize}

Once again, it should be pointed out that nowhere in the network there
is such a global vision of these exection phases. A more detailed
description of the exection model at each node is discussed in the
next section.

\subsection{Packets types Required}

@@

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the DiSR algorithm.
The node is not marked as \emph{visited}/\emph{tvisited}.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
% TODO: this and all the other details on the items below should be
% checked and moved elsewhere for a future long paper

%This is required because a
%\texttt{STARTING\_SEGMENT\_REQUEST} packet should be injected in order to start
%the whole process.
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
%This happens when all
%the following conditions are matched: the node is marked as \emph{visited},
%that is \emph{visited} = TRUE 
%TODO: \emph{segID} is set to some value reflecting
%the segment request sent there is a link that is being investigated by
%the node, that is a tvisited\_link marked with the same id of the node 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
with id segid, not being itself the node from which the find process
was started.

%The following conditions should be matched: \emph{tvisited} =
%true, with segID=X different from node's id.  Two links set as
%\emph{link\_tvisited} with id X. Let’s refer them as  links 'i' and 'j'. Then
%link\_tvisited[i] was set = X when the find process from an adjacent
%node reached the current node, while link\_tvisited[j] has been set to
%X when the node itself started to investigate its free links for the
%segment request with id X. Note that while investigating its attached
%links, if the find process fails along path 'j', the current node can
%investigate another free node 'k' if suitable.  
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment.

%Two main differences: Can confirm
%\texttt{STARTING\_SEGMENT\_CONFIRM} with the same \emph{segID} when receiving a normal
%\texttt{SEGMENT\_REQUEST}: Overwrites its status setting it to \emph{Candidate} for the
%segID frees the links previously set as \emph{tvisited} during flooding
\item{\emph{Assigned}}: a node for which the segment has been determined.  The
segment \emph{segID} attribute value is set to some id X different from \texttt{NULL}.
%The \emph{visited} value is \emph{true} NOTE: an \emph{Assigned} state is a quick temporary
%state, since it becomes \emph{ActiveSearching} if a not (\emph{visited}/tvisited)
%link is suitable for searching new segments (see next\_visited
%procedure on paper) 
\end{itemize}

\subsection{Sketch of DiSR}

\begin{itemize}

\item{\textbf{Injecting bootstrap request}}: all nodes have a initial
\emph{DBS} status \emph{Free}, except for a
 node with status \emph{Bootstrap}, set by some signal from an
upper layer via. When starting, this bootstrap node sets itself as \emph{visited} and
changes its status to \emph{ActiveSearching}, injecting a
\texttt{STARTING\_SEGMENT\_REQUEST} across one of its free links.

\item{\textbf{Flooding}}: Each node receiving a \texttt{STARTING\_SEGMENT\_REQUEST},
if not yet (\emph{visited}/\emph{tvisited}),  forwards it to its free links using a flooding mechanism.
Each of these links is then marked as \emph{tvisited} with the segment id
associated to the request. Note that a node that has already received a
\texttt{STARTING\_SEGMENT\_REQUEST} packet can simply ignore further packets
associated to the same request, having already contributed to the
flooding. 
%Note also that each of these packets has a max\_segment\_hops
%field to prevent packets from travelling undefinitively.

\item{\textbf{Confirming the Starting Segment}}: when this \texttt{STARTING\_SEGMENT\_REQUEST} packet reaches
the starting node (from a different link, of course), the starting
segment is found. Then the starting node sends back a \texttt{STARTING\_SEGMENT\_CONFIRM}
packet with the same id along the link from which it received the
\texttt{STARTING\_SEGMENT\_REQUEST}. Each node do same by changing its own status
to \emph{Assigned}. So the confirmation packet is sent back from
node to node and the starting segment is created.

Note: A node knows to be \emph{Candidate} due a
\texttt{STARTING\_SEGMENT\_REQUEST}.  Then, when receiving a simple
(not starting) \texttt{SEGMENT\_REQUEST} , it can simply cancel its
previous \emph{Candidate} status, since this second
request means that the starting segment has already been found.

@@

\item{\textbf{Injecting other request}}: Each node in the \emph{Assigned} status can initiate a search for a
segment, by sending a \texttt{SEGMENT\_REQUEST} across one of its free
links. Note that, since this is not the starting segment, in this case
the packet should not return back to the initiator, but just another visited
node
\item{\textbf{Setup of a segment}}: The find process is successfull
when an \emph{Assigned} node receives the \texttt{SEGMENT\_REQUEST} packet. Then, a \texttt{SEGMENT\_CONFIRM} is
sent back along the path that originated the request.  
\item{\textbf{Failing while searching a segment}}: a node received a \texttt{SEGMENT\_REQUEST} packet but
matched one of the following conditions: the node is free but has no
more suitable free links (can’t forward the \texttt{SEGMENT\_REQUEST}) the node
is \emph{tvisited} candidate for another find process the request packet
exceeded the max\_segment\_hops limit.  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the proper link modifying its state
to free. Each of the subsequent nodes will forward this \texttt{SEGMENT\_CANCEL}
only if no more free links can be explored starting from them.
\end{itemize}

%\subsection{Intra-node vs Inter-node parallelism}
%In the processes described above, we assumed that each node in the
%READY\_SEARCHING state can start a find segment process by injecting a
%\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when defining DiSR
%should be made in order to decide whether or not the node must
%actually perform this action. 
%The point above is is strictly related to the general question: which
%level of “parallelism” should we allow in DiSR ? The adopted approach
%is that, although the nature of DiSR itself is intrinsicly parallel,
%the use of parallelism makes things work in a more complex way. In
%other words, DiSR is parallel when needed, but it does not exploit
%parallelism as an “improving feature”. When not needed, things should
%be serialized. For example: when a node is “ready searching” could
%start several find segment process associated to the same segment, one
%for each free link. But serializing this by investigating the free
%links in order could be a simplest solution. We refer to this saying
%that we avoid intra-node parallelism.  Note that while intra-node
%parallelism can be avoided, trying to avoid inter-node parallelism
%could be more complex than allowing the parallelism itself. Imagine
%for example the effort of trying to coordinate nodes so that an unique
%finding node process is actually running in the whole subnet. Thus, in
%contrast to the intra-node parallelism, the inter-node parallelism is
%a structural property of the DiSR algorithm and should not be avoided.
%------------------------------------------------------------------------------

\section{Detailed Node behaviour Model}
\label{sec:disr_detailed_model}

@@

\subsection{Receiving a \texttt{SEGMENT\_CONFIRM}}
When received, if the node status is \emph{Candidate} with a \emph{segID} corresponding to the one indicated in
the packet, the node learns to belong to the segment associated to the
id. Further, it should forward this packet to the link where the
original \texttt{SEGMENT\_REQUEST} packet came from, so that all candidate nodes
can learn the segment id they belong to.  \emph{LED} shoud be
updated: The state of the node changes from \emph{tvisited} to \emph{visited}.  the
\emph{link\_tvisited} previously set with the \emph{segID} should be invalidated the
\emph{link\_visited} variable associated to the incoming confirm should be set
as segID. Note that if the request was flooded across different links,
the \emph{link\_visited} variable associated to the other paths should not be
set.  the \emph{link\_visited} variable associated to the link from which the
request was originated should be set segID

>>>>

\subsection{Basic Idea}
The main concept behind the kind of turn prohibition we want to
achieve is the \emph{Segment}: it is basically a path of consecutive
nodes and links. A segment $S_1$ starts with a link attached to a node which
belongs to different segment $S_2$, ending with a link attached to a
node belonging to a different segment $S_3$. In other words, a segment
a path between connecting two other different segments.A exception is
the first segment established in the network, called \emph{starting
segment} which is  a loop beginning/ending on a particular node
defined as \emph{bootstrap node}.
The idea is to partition the network in a set of disjoint segments, and then
placing a turn restriction within each segment. It has been proved~\cite{mejia_ipdps06}
that such a set of turn prohibitions guarantees deadlock freedom and
while preserving connectivity of the network.

@@

Roughly speaking, the execution of can be described as the execution
of the following phases: (1) Injection of the DiSR process from upper layer to set a bootstrap
node (2) Bootstrap node broadcasting to create the first segment of the subnet
(3) Parallel requests starting from assigned node to discover other
segments. The following subsection describe the data structure and
execution model required at each node. (4) Turn proibition in each
confirmed segment.
%\item Draining of remaining packet due timeouts

%Once again, it should be pointed out that nowhere in the network there
%is such a global vision of these exection phases. A more detailed
%description of the exection model at each node is discussed in the
%next section.

\subsection{Signal Types Required}

@@

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the DiSR algorithm.
The node is not marked as \emph{visited}/\emph{tvisited}.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
% TODO: this and all the other details on the items below should be
% checked and moved elsewhere for a future long paper
%This is required because a
%\texttt{STARTING\_SEGMENT\_REQUEST} packet should be injected in order to start
%the whole process.
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
%This happens when all
%the following conditions are matched: the node is marked as \emph{visited},
%that is \emph{visited} = TRUE 
%TODO: \emph{segID} is set to some value reflecting
%the segment request sent there is a link that is being investigated by
%the node, that is a tvisited\_link marked with the same id of the node 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
with id segid, not being itself the node from which the find process
was started.

%The following conditions should be matched: \emph{tvisited} =
%true, with segID=X different from node's id.  Two links set as
%\emph{link\_tvisited} with id X. Let’s refer them as  links 'i' and 'j'. Then
%link\_tvisited[i] was set = X when the find process from an adjacent
%node reached the current node, while link\_tvisited[j] has been set to
%X when the node itself started to investigate its free links for the
%segment request with id X. Note that while investigating its attached
%links, if the find process fails along path 'j', the current node can
%investigate another free node 'k' if suitable.  
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment. 
%Two main differences: Can confirm
%\texttt{STARTING\_SEGMENT\_CONFIRM} with the same \emph{segID} when receiving a normal
%\texttt{SEGMENT\_REQUEST}: Overwrites its status setting it to \emph{Candidate} for the
%segID frees the links previously set as \emph{tvisited} during flooding
\item{\emph{Assigned}}: a node for which the segment has been determined.  The
segment \emph{segID} attribute value is set to some id X different from \texttt{NULL}.
%The \emph{visited} value is \emph{true} NOTE: an \emph{Assigned} state is a quick temporary
%state, since it becomes \emph{ActiveSearching} if a not (\emph{visited}/tvisited)
%link is suitable for searching new segments (see next\_visited
%procedure on paper) 
\end{itemize}

\subsection{DiSR Execution Model}

In Figure~\ref{fig:state_machine} are depicted the main events
representing the execution model at each node. The DiSR control
packets described in the previous subsection trigger node events,
which eventually change $DBS$ status according to the current value of
$LED$ variables and the type of the packet received. The main DiSR
phases, together with a reference to the particular status transition
shown in Figure~\ref{fig:state_machine} are described in the
following.

\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/state_machine.eps}
  \caption{DiSR node execution model}
  \label{fig:state_machine}
\end{figure}
\begin{itemize}

\item{\textbf{Injecting bootstrap request}}: all nodes have an initial
\emph{DBS} status \emph{Free}, except for a
 node with status \emph{Bootstrap}, set by some signal from an
upper layer via (a). When starting, this bootstrap node 
changes its status to \emph{ActiveSearching} (b), injecting a
\texttt{STARTING\_SEGMENT\_REQUEST} across one of its free links.

\item{\textbf{Flooding}}: Each node receiving a \texttt{STARTING\_SEGMENT\_REQUEST},
when $Free$ forwards it to its free links using a flooding mechanism and
becoming $CandidateStarting$ (g).
Each of its free links is then marked as \emph{tvisited} with the segment id
associated to the request. Note that a node that has already received a
\texttt{STARTING\_SEGMENT\_REQUEST} packet can simply ignore further packets
associated to the same request, having already contributed to the
flooding. 
%Note also that each of these packets has a max\_segment\_hops
%field to prevent packets from travelling undefinitively.

\item{\textbf{Confirming the Starting Segment}}: when the \texttt{STARTING\_SEGMENT\_REQUEST} packet reaches
the bootstrap node (from a different link, of course), the starting
segment is found. Then the bootstrap node sends back a \texttt{STARTING\_SEGMENT\_CONFIRM}
packet along the link from which it received the
\texttt{STARTING\_SEGMENT\_REQUEST} and becomes $Assigned$ (c). Each
node receiving the confirm do same by changing its own status
to \emph{Assigned} (l). So the confirmation packet is sent back from
node to node and the starting segment is created.

Note: A node being \emph{CandidateStarting}, when receiving a simple
(not starting) \texttt{SEGMENT\_REQUEST} , can simply cancel its
previous \emph{CandidateStarting} status and set itself as
$Candidate$ for that request (h), since this new
request means that the starting segment has already been found.

@@

\item{\textbf{Injecting other requests}}: Each node in the
\emph{Assigned} status can initiate a search for a segment, by sending
a \texttt{SEGMENT\_REQUEST} across one of its free links (d). Note
that since this is not the starting segment in this case the packet
should not reach the initiator, but just another \emph{Assigned} node.

\item{\textbf{Setup of a segment}}: The find process is successful
when an already \emph{Assigned} node receives the
\texttt{SEGMENT\_REQUEST} packet. Then, a \texttt{SEGMENT\_CONFIRM} is
sent back along the path that originated the request (e), while the
node remain $Assigned$ (since it could confirm more segments). Each node
previously set as $Candidate$ for that segment id, when receiving the
confirm packet changes its status to $Assigned$ (m) and forwards back
the same confirm until it reaches the initiator of the request, which
changes from $ActiveSearching$ to $Assigned$ (c).

\item{\textbf{Failing while searching a segment}}: a node received a \texttt{SEGMENT\_REQUEST} packet but
matched one of two the following conditions: the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})(f); the node
is candidate for another find process(i).  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the proper link. 
%Each of the subsequent nodes will forward this \texttt{SEGMENT\_CANCEL}
%only if no more free links can be explored starting from them.
\end{itemize}

\subsection{Intra-node vs Inter-node parallelism}
In the processes described above, we assumed that each node in the
READY\_SEARCHING state can start a find segment process by injecting a
\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when defining DiSR
should be made in order to decide whether or not the node must
actually perform this action. 
The point above is is strictly related to the general question: which
level of “parallelism” should we allow in DiSR ? The adopted approach
is that, although the nature of DiSR itself is intrinsicly parallel,
the use of parallelism makes things work in a more complex way. In
other words, DiSR is parallel when needed, but it does not exploit
parallelism as an “improving feature”. When not needed, things should
be serialized. For example: when a node is “ready searching” could
start several find segment process associated to the same segment, one
for each free link. But serializing this by investigating the free
links in order could be a simplest solution. We refer to this saying
that we avoid intra-node parallelism.  Note that while intra-node
parallelism can be avoided, trying to avoid inter-node parallelism
could be more complex than allowing the parallelism itself. Imagine
for example the effort of trying to coordinate nodes so that an unique
finding node process is actually running in the whole subnet. Thus, in
contrast to the intra-node parallelism, the inter-node parallelism is
a structural property of the DiSR algorithm and should not be avoided.
%------------------------------------------------------------------------------

% TODO: FIX and RE-enable in long paper

\section{Detailed Node behaviour Model}
\label{sec:execution_model}

@@

\subsection{Receiving a \texttt{SEGMENT\_CONFIRM}}
When received, if the node status is \emph{Candidate} with a \emph{segID} corresponding to the one indicated in
the packet, the node learns to belong to the segment associated to the
id. Further, it should forward this packet to the link where the
original \texttt{SEGMENT\_REQUEST} packet came from, so that all candidate nodes
can learn the segment id they belong to.  \emph{LED} shoud be
updated from \emph{tvisited} to \emph{visited}.  the
\emph{link\_tvisited} previously set with the \emph{segID} should be invalidated the
\emph{link\_visited} variable associated to the incoming confirm should be set
as segID. Note that if the request was flooded across different links,
the \emph{link\_visited} variable associated to the other paths should not be
set.  the \emph{link\_visited} variable associated to the link from which the
request was originated should be set segID
